Fast Fourier transformation of electromagnetic data for computationally expensive kernels

### 2021-04-07 ManuscriptTime.tex
Revision II marked changes

\documentclass[extra, camera,%
    onecolumn,   % comment for two columns
    referee,     % uncomment for double lines
    final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier
transformation, and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time.

\added{Electromagnetic methods in geophysics span a wide range of model
scenarios and acquisition layouts which have all different implications. A
well-known case is for instance the land or shallow marine case versus the deep
marine case, where in the former case a dominating airwave has to be accurately
modelled which can can be completely ignored in the latter case. Recent areas
of particular interest in time-domain modelling that pose numerical challenges
are, for instance, simulating the fields through steel-cased wells
\mbox{\citep{CAG.19.Heagy}} or the effects of induced polarization
\mbox{\citep{EXP.20.Kang}}. The former is challenging because of the high
conductivity contrasts requiring very detailed meshing, the latter is
challenging for time-domain modelling because the models are frequency
dependent.}

@@

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity \added{of different time-domain methods
with the computational complexity of frequency-domain computation followed by a
Fourier transform:} \deleted{of} explicit time-stepping schemes (Du
Fort-Frankel method as used in, e.g., \cite{GEO.04.Commer,
GEO.07.Maao})\replaced{ and}{;} implicit schemes \citep[e.g.,
][]{GEO.04.Haber}\added{ have a complexity of $O(n^4)$}, \deleted{and} matrix
exponentials and Lanczos reduction schemes \cite[e.g., ][]{RS.94.Druskin}
\added{have a complexity of $O(n^4\sqrt{\log n})$}\replaced{, and}{with the}
Fourier transform of frequency-domain solutions \added{have a complexity $O(n_f
n^3)$}; \added{$n$ is the number of grid points in each direction, and $n_f$
the number of frequencies}. Their conclusion was that the Fourier transform
method can be favourable if the number of required frequencies is small
relative to the number of grid points in each coordinate. The conditions for
fast computation of time-domain data with a frequency-domain code are: a
sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are \added{the further reduction of required
frequencies by introducing lower and upper thresholds of numerically important
frequencies,} an adaptive, frequency-dependent gridding scheme that minimizes
the required cells in each dimension, and a logarithmic Fourier transform such
as digital linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast
Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency
to the time domain. The latter makes it also possible to only use the imaginary
part of the frequency-domain response, which has advantages when it comes to
interpolation.

While we use the finite-integration technique \citep{AEU.77.Weiland} in the
frequency domain to demonstrate our method, it is important to note that our
ideas can be used with any solver and is not bound to one or another frequency-
or time-domain solver. The ideas for the frequency selection can be applied to
time selection, and time-dependent, adaptive gridding exists as well
\citep[e.g., ][]{RS.06.Commer}. In the next section, we briefly review the
methodology as introduced by \cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and
highlight their advantages and shortcomings. This is followed by an outline of
our changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains.
Another approach is to have Maxwell's equations only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the
frequency domain to the time domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies\replaced{ through a combination of
extrapolation, interpolation, and setting the responses for high frequencies to
zero. This translates}{, translating} into a significant reduction in
computation time. \replaced{The number of required frequencies is not the only
important point for reducing computation time. Another important aspect is the computation grid, as t}{T}he required \deleted{computation} grids
for low frequencies (in our case in the order of 0.001\,Hz) and high
frequencies (around 100\,Hz) are hugely different. Low frequencies can be
computed on a coarser grid, but they require a much larger domain in order to
avoid boundary effects. High frequencies, on the other hand, require denser
gridding, but they are much more limited in reach. An adaptive gridding scheme
is therefore required\replaced{, which}{. We build our approach upon
\mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field is decayed by $1/e\approx 37\,\%$. The skin depth
$\delta$ is a function of conductivity and frequency, and for the diffusive
approximation of Maxwell's equations in an isotropic, homogeneous medium is
given by \citep[e.g.,][\mbox{\deleted{ equation~1.53}}]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes. It is
important to note that while we show the transform from frequency to time
domain, the conclusions regarding the Fourier transform can equally well be
applied to the transform from time-domain to frequency-domain data for a code
that solves Maxwell's equations in the time domain. \added{Our recommendations
for speeding up the Fourier transform for expensive kernels are independent of
spatial complexity, as diffusive EM responses are smooth functions of both
frequency and time. However, the intrinsic requirement is an accurate solver
over a sufficient wide range of frequencies or times.} \deleted{A variation of
our method was implemented, for instance, by \custem
\mbox{\citep{GEO.19.Rochlitz}} to obtain frequency-domain data from time-domain
computations (Rochlitz, private communication, 2020).}

@@

\added{The requirement for any transform is a robust solver of sufficient
precision over a wide range of frequencies.} \replaced{We use f}{F}or the
numerical computations \deleted{we use} the open-source (Apache License 2.0)
code\added{s \empymod} \added{\mbox{\citep{GEO.17.Werthmuller}}} \added{and}
\emg3d \citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a \added{3D}
multigrid solver based on \cite{GP.06.Mulder}, which can be used as a
preconditioner for Krylov subspace solvers or as a solver on its own. The
multigrid approach works fine for the diffusive approximation of Maxwell's
equations, which assumes that $\omega\varepsilon \ll \sigma$, where
$\varepsilon$ is electric permittivity (F/m). The remaining system to solve in
the frequency domain is then given by the second-order differential equation of
the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All \deleted{the} examples in this article are run on a
laptop with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using
Ubuntu 20.04 and Python 3.8. \replaced{They}{All examples} were run using a
single thread, but parallelization over frequencies is straight forward, as the
computations are independent. \deleted{For the semi-analytical computations of
layered models we use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

\added{Note that while we use these two algorithms for the numerical examples
any solver that provides sufficient accurate and precise frequency-domain
responses could be used.}

@@

\replaced{Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required frequencies
for the transform, but to minimize the number of frequencies for which we
actually have to compute responses. We do this in three zones: Above an upper
threshold $f_\mr{max}$ we set the frequency-domain response to zero. Below a
lower threshold $f_\mr{min}$ we extrapolate the frequency-domain response. As
we know that the imaginary response goes towards zero for zero frequency, we
can turn it into an interpolation, for which we use PCHIP. And in-between the
thresholds we use cubic spline interpolation in order to only have to actually
compute a few distinct frequencies per decade ($n_\mr{dec}^f$). Using this
approach we found that computing the responses for 15--25 frequencies is
usually good enough to obtain time-domain responses for a large range of
offsets; even though the actual transform might need up to hundreds of
frequencies.}{There are three key-parameters that have to be defined: Minimum
and maximum frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies
per decade ($n_\mr{dec}^f$).}

@@

\added{The proposed Fourier transformation requires a robust solver that can
compute accurate results over a wide range of frequencies (times) to obtain
time-domain (frequency-domain) responses. In our case the solver uses a
regular, stretched grid and computes the electric fields in the frequency
domain. This setup is the target of the following gridding recommendations.
Whilst they will look differently for other mesh types or a time-domain code,
some conclusion will still hold.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the
\deleted{perfectly electrically conducting} boundary condition. It can also
have a much higher stretching factor $\alpha_\mr{c}$. In the presented examples
we have chosen $D_\mr{c}$ such that the distance for the signal diffusing from
the source to the boundary and back to the receiver closest to the boundary is
at least two wavelengths, after which the amplitude of the signal is reduced to
a millionth of its initial strength. The wavelength $\lambda$ to compute
$D_\mr{c}$ is given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth.
\deleted{Note that for shallow marine and land cases t}\added{T}his also
applies to the horizontal dimensions, not only to the upward
$z$-direction\added{,} and \replaced{it equally applies}{similarly} for very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here. \added{Note that grid stretching
for complex-valued diffusion fields is essentially what is done for wave fields
with the perfectly matched layer (PML). PML is an absorbing layer to avoid
scattering from the boundary by letting the field die out to zero, which is
achieved by introducing a complex factor which causes damping. As
electromagnetic diffusion fields are damped fields by themselves it suffices to
stretch the grid.}

@@

\added{The proposed Fourier-transform approach works independently of spatial
complexity, as diffusive electromagnetic fields are smooth functions of both
frequency and time. An interesting test it to see whether this applies as
well for dispersive media where the model parameters are frequency dependent,
such as is the case in induced polarization. To test this we use the Cole-Cole
model \mbox{\citep[CCM, ][]{JCP.41.Cole}} which, written in terms of
conductivities instead of electric permittivities as by the original authors,
is given by}
%
\begin{equation}
\added{
  \sigma(\omega) = \sigma_\infty + \frac{\sigma_0 - \sigma_\infty}{1 +
    (\rm{i}\omega\tau)^c}\ ,
}
  \label{eq:ccm}
\end{equation}
%
\added{where $\sigma_0$ and $\sigma_\infty$ refer to the low-frequency and
high-frequency conductivity values, respectively, $\tau$ is the central
relaxation time (s), and $c$ the CCM exponent describing the broadness of the
relaxation time distribution. Note that this model slightly differs from the
one phrased in terms of resistivities given by \mbox{\cite{GEO.78.Pelton}},
see, e.g., \mbox{\cite{GJI.13.Tarasov}}. }

>>>>

\documentclass[extra, camera,%
    onecolumn,   % comment for two columns
    referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
Three-dimensional controlled-source electromagnetic data is often computed
directly in the domain of interest, either in the frequency domain or in the
time domain. Computing it in one domain and transforming it via a Fourier
transform to the other domain is a viable alternative. It requires the
evaluation of many responses in the computational domain if standard Fourier
transforms are used. This can make it prohibitively expensive if the kernel is
time-consuming as is the case in three-dimensional electromagnetic modelling.
The speed of such modelling that is obtained through a transform is defined by
three key points: solver, method and implementation of the Fourier
transform\deleted{ation}, and gridding.
%
The faster the solver, the faster modelling will be. It is important that the
solver is robust over a wide range of values (frequencies or times). The method
should require as few kernel evaluations as possible while remaining robust. As
the frequency and time ranges span many orders of magnitude, the required
values are ideally equally spaced on a logarithmic scale. The proposed fast
method uses either the digital linear filter method or the logarithmic fast
Fourier transform together with a careful selection of evaluation points and
interpolation. In frequency-to-time domain tests this methodology requires
typically 15 to 20 frequencies to cover a wide range of offsets.
%
The gridding should be frequency- or time-dependent, which is accomplished by
making it a function of skin depth. Optimising for the least number of required
cells should be combined with optimising for computational speed. Looking
carefully at these points results in much smaller computation times with
speedup factors of ten or more over previous methods. A computation in one
domain followed by transformation can therefore be an alternative to
computation in the other domain domain if the required evaluation points and
the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or for geological purposes \citep{NAT.19.Johanson}.
While current sources with a few frequencies are used in the deep marine
environment, transient measurements are more common in the shallow marine
environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis,
SEG.07.Avdeeva}. One of the main reasons is the dominance of the airwave in
shallow marine and terrestrial measurements, which can be better separated in
the time domain. CSEM is usually divided into frequency- and time-domain
methods, depending on whether the source signal is a continuous waveform, such
as a sine, or a finite waveform, such as a pseudo-random binary sequence
(PRBS). A numerical comparison of the two methods is given by
\cite{GP.13.Conell}. Acquired CSEM data are subsequently often analysed
(modelled and inverted for) in their respective domain, either the frequency or
time domain. Modellers of layered media usually exploit the horizontal
shift-invariance by computing the responses in the wavenumber-frequency domain
followed by a 2D inverse spatial Fourier transform, also called Hankel
transform, to the space-frequency domain, and a regular inverse Fourier
transform if time-domain data are required \citep[e.g., ][]{GEO.15.Hunziker}.
CSEM codes for 2D and 3D computations, on the other hand,
\replaced{often}{mostly} compute their responses directly in the required
domain, either frequency or time.

\added{Electromagnetic methods in geophysics span a wide range of model
scenarios and acquisition layouts, each with its own modelling-related
implications. A well-known example is the aforementioned land or shallow marine
case versus the deep marine case, where in the former case a dominating airwave
has to be accurately modelled, whereas it can be completely ignored in the
latter case. Recent areas of particular interest in time-domain modelling that
pose numerical challenges are, for instance, simulating the fields through
steel-cased wells \mbox{\citep{CAG.19.Heagy}} or the effects of induced
polarization \mbox{\citep{EXP.20.Kang}}. The former is challenging because of
the high conductivity contrasts requiring very detailed meshing, the latter is
challenging for time-domain modelling because the models are frequency
dependent.}

@@

Under certain conditions it can be a viable alternative to model time-domain
data with a frequency-domain code, as shown by \cite{GEO.08.Mulder}. They
compared the computational complexity \added{of different time-domain methods
with the computational complexity of frequency-domain computation followed by a
Fourier transform:} \deleted{of} explicit time-stepping schemes (Du
Fort-Frankel method as used in, e.g., \cite{GEO.04.Commer,
GEO.07.Maao})\replaced{ and}{;} implicit schemes (e.g.,
\cite{GEO.04.Haber}\added{; \mbox{\cite{GEO.10.Um}}}) \added{ have a complexity
of $O(n^4)$}, \deleted{and} matrix exponentials and Lanczos reduction schemes
\cite[e.g., ][]{RS.94.Druskin} \added{have a complexity of $O(n^4\sqrt{\log
n})$}\replaced{, and}{with the} Fourier transform of frequency-domain solutions
\added{have a complexity $O(n_f n^3)$}; \added{$n$ is the number of grid points
in each direction, and $n_f$ the number of frequencies}. Their conclusion was
that the Fourier transform method can be favourable if the number of required
frequencies is small relative to the number of grid points in each coordinate.
The conditions for fast computation of time-domain data with a frequency-domain
code are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are \added{the further reduction of required
frequencies by introducing lower and upper thresholds of numerically important
frequencies,} an adaptive, frequency-dependent gridding scheme that minimizes
the required cells in each dimension, and a logarithmic Fourier transform such
as digital linear filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast
Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency
to the time domain. The latter makes it also possible to only use the imaginary
part of the frequency-domain response, which has advantages when it comes to
interpolation.

\deleted{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to demonstrate our
method, it is important to note that our ideas can be used with any solver and
is not bound to one or another frequency- or time-domain solver. The ideas for
the frequency selection can be applied to time selection, and time-dependent,
adaptive gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}.}} In the
next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of \added{the methodology,}
our changes to the Fourier transform and the adaptive gridding. Finally, we
demonstrate the efficiency of the approach with some numerical results.
\added{These compromise of a homogeneous space, a layered model, a model which
includes dispersive media (induced polarization), and a three-dimensional
model.}

@@

Being able to model CSEM data both in the frequency domain and in the time
domain can be desirable, as both domains have advantages and disadvantages. One
way to achieve this is to implement Maxwell's equations in both domains
\added{as is done, for instance, in \texttt{SimPEG}
\mbox{\citep{CAG.15.Cockett}} or in \texttt{custEM}
\mbox{\citep{GJI.21.Rochlitz}}}. Another approach is to have Maxwell's equations
only implemented in one domain, and use Fourier transforms to go to the other.
However, this approach can be costly, as many frequencies over a wide range are
required to go from the frequency domain to the time domain, or many times over
a wide range for the opposite direction. We present a methodology which
significantly reduces the amount and range of the required
frequencies\replaced{ through a combination of extrapolation, interpolation,
and setting the responses for high frequencies to zero. This translates}{,
translating} into a significant reduction in computation time. \replaced{The
number of required frequencies is not the only important point for reducing
computation time. Another important aspect is the computation grid, as t}{T}he
required \deleted{computation} grids for low frequencies (in our case in the
order of 0.001\,Hz) and high frequencies (around 100\,Hz) are hugely different.
Low frequencies can be computed on a coarser grid, but they require a much
larger domain in order to avoid boundary effects. High frequencies, on the
other hand, require denser gridding, but they are much more limited in reach.
An adaptive gridding scheme is therefore required\replaced{, which}{. We build
our approach upon \mbox{\cite{GEO.07.Plessix}}, who presented such an adaptive
gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain} is naturally based on the skin depth,
\deleted{which is} the distance after which the amplitude of the
electromagnetic field \replaced{has}{is} decayed by $1/e\approx 37\,\%$. The
skin depth $\delta$ is a function of conductivity and frequency, and for the
diffusive approximation of Maxwell's equations in an isotropic, homogeneous
medium is given by \citep[e.g.,][\mbox{\deleted{
equation~1.53}}]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \ 503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain \added{using the finite-integration technique
\mbox{\citep{AEU.77.Weiland}}}, which makes time-domain modelling with a
frequency-domain code viable in comparison with time-domain codes.
It is important to note that \added{our ideas can be used with any solver and
is not bound to one or another frequency- or time-domain solver. The ideas for
the frequency selection can be applied to time selection, and time-dependent,
adaptive gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}.}}
\deleted{while we show the transform from frequency to time domain, the
conclusions regarding the Fourier transform can equally well be applied to the
transform from time-domain to frequency-domain data for a code that solves
Maxwell's equations in the time domain.} \added{Our recommendations for
speeding up the Fourier transform for expensive kernels are independent of
spatial complexity, as diffusive EM responses are smooth functions of both
frequency and time. However, the intrinsic requirement is an accurate solver
over a sufficient wide range of frequencies or times.} \deleted{A variation of
our method was implemented, for instance, by \custem
\mbox{\citep{GEO.19.Rochlitz}} to obtain frequency-domain data from time-domain
computations (Rochlitz, private communication, 2020).}

@@

\added{The requirement for any transform is a robust solver of sufficient
accuracy over a wide range of frequencies.} \replaced{We use f}{F}or the
numerical computations \deleted{we use} the open-source (Apache License 2.0)
code\added{s \empymod} \added{\mbox{\citep{GEO.17.Werthmuller}}} \added{and}
\emg3d \citep{JOSS.19.Werthmuller}\replaced{.}{,} \added{The former computes
semi-analytical responses of layered models, the latter is} a \added{3D}
multigrid solver based on \cite{GP.06.Mulder}, which can be used as a
preconditioner for Krylov subspace solvers or as a solver on its own. The
multigrid approach works fine for the diffusive approximation of Maxwell's
equations, which assumes that $\omega\varepsilon \ll \sigma$, where
$\varepsilon$ is electric permittivity (F/m). The remaining system to solve in
the frequency domain is then given by the second-order differential equation of
the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which known improvements such as line-relaxation and semicoarsening
\citep{ECCFD.06.Jonsthovel} are implemented, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All \deleted{the} examples in this article are run on a
laptop with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using
Ubuntu 20.04 and Python 3.8. \replaced{They}{All examples} were run using a
single thread, but parallelization over frequencies is straight forward, as the
computations are independent. \deleted{For the semi-analytical computations of
layered models we use the open-source (Apache License 2.0) code \empymod
\mbox{\citep{GEO.17.Werthmuller}}.}

\added{Note that while we use these two algorithms for the numerical examples,
any solver that provides sufficiently accurate frequency-domain responses could
be used.}

@@

\replaced{Any Fourier transform needs a certain range of frequencies to obtain
time-domain results. Our approach is not to minimize the required frequencies
for the transform, but to minimize the number of frequencies for which we
actually have to compute responses. We do this in three zones: above an upper
threshold $f_\mr{max}$, we set the frequency-domain response to zero. Below a
lower threshold $f_\mr{min}$, we extrapolate the frequency-domain response. As
we know that the imaginary response goes towards zero for zero frequency, we
can turn it into an interpolation, for which we use PCHIP. And in-between the
thresholds, we use cubic spline interpolation in order to only have to actually
compute a few distinct frequencies per decade ($n_\mr{dec}^f$). Using this
approach, we found that computing the responses for 15--25 frequencies is
usually good enough to obtain time-domain responses for a large range of
offsets; even though the actual transform might need up to hundreds of
frequencies.}{There are three key-parameters that have to be defined: Minimum
and maximum frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies
per decade ($n_\mr{dec}^f$).}

@@

\added{The proposed Fourier transform requires a robust solver that can compute
accurate results over a wide range of frequencies (times) to obtain time-domain
(frequency-domain) responses. In our case the solver uses a regular, stretched
grid and computes the electric fields in the frequency domain. This setup is
the target of the following gridding recommendations. Whilst they will look
differently for other mesh types or a time-domain code, some conclusion will
still hold.}

The computation grid consists of a core or survey domain $D_\mr{s}$ that should
contain all source and receiver positions. The survey domain usually has no or
very little cell stretching factor $\alpha_\mr{s}$. The minimum cell width is
defined as
%
\begin{equation}
  \Delta_\mr{min}=\delta(f, \sigma_\mr{src})/n_\delta \, ,
  \label{eq:minwidth}
\end{equation}
%
where $\sigma_\mr{src}$ is the conductivity of the media in which the source
resides, and $n_\delta$ is a positive number that defines how many cells there
should be per skin depth. The actual computational domain $D_\mr{c}$ is usually
much bigger than $D_\mr{s}$ in order to avoid artefacts from the
\deleted{perfectly electrically conducting} boundary condition. It can also
have a much higher stretching factor $\alpha_\mr{c}$. In the presented examples
we have chosen $D_\mr{c}$ such that the distance for the signal diffusing from
the source to the boundary and back to the receiver closest to the boundary is
at least two wavelengths, after which the amplitude of the signal is reduced to
a millionth of its initial strength. The wavelength $\lambda$ to compute
$D_\mr{c}$ is given by
%
\begin{equation}
  \lambda = 2\pi\delta(f, \sigma_\mr{ave})
  \ \approx \ 3162/\sqrt{f\sigma_\mr{ave}}\, ,
 \label{eq:lambda}
\end{equation}
%
where $\sigma_\mr{ave}$ is the average conductivity, which can vary for
different directions. However, the skin-depth approach fails for air, in which
the EM field propagates as a wave at the speed of light. A largest
computational domain is therefore enforced, defining the maximum distance from
the source to the boundary; this distance is by default set to 100\,km, but
this can be reduced in the marine case with increasing water depth.
\deleted{Note that for shallow marine and land cases t}\added{T}his also
applies to the horizontal dimensions, not only to the upward
$z$-direction\added{,} and \replaced{it equally applies to}{similarly for} very
resistive basements, even in deep water. One way to circumvent this difficulty
is the use of a primary-secondary formulation, where the primary field,
including the air wave, is computed with a semi-analytical code for layered
media. We do not consider this approach here. \added{Note that grid stretching
for complex-valued diffusion fields is essentially what is done for wave fields
with the perfectly matched layer (PML). PML is an absorbing layer to avoid
scattering from the boundary by letting the field decay to zero, which is
achieved by introducing a complex factor that causes damping. As
electromagnetic diffusion fields are damped fields by themselves, it suffices
to stretch the grid.}

@@

\added{The proposed Fourier-transform approach works independently of spatial
complexity, as diffusive electromagnetic fields are smooth functions of both
frequency and time. An interesting test it to see whether this applies as
well for dispersive media where the model parameters depend on frequency, as is
the case in induced polarization. To test this we use the Cole-Cole model
\mbox{\citep[CCM, ][]{JCP.41.Cole}} which, written in terms of conductivities
instead of electric permittivities as by the original authors, is given by}
%
\begin{equation}
\added{
  \sigma(\omega) = \sigma_\infty + \frac{\sigma_0 - \sigma_\infty}{1 +
    (\rm{i}\omega\tau)^c}\ ,
}
  \label{eq:ccm}
\end{equation}
%
\added{where $\sigma_0$ and $\sigma_\infty$ refer to the low-frequency and
high-frequency conductivity values, respectively, $\tau$ is the central
relaxation time (s), and $c$ the CCM exponent describing the broadness of the
relaxation time distribution. Note that this model slightly differs from the
one phrased in terms of resistivities given by \mbox{\cite{GEO.78.Pelton}},
see, e.g., \mbox{\cite{GJI.13.Tarasov}}. }
